Open-source tools in R for landscape ecology \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-12-21 paper/paper.Rmd
:pencil: Address some comments

<!-- JN: my first thought was - "do we even need this kind of introduction in the 'Current Landscape Ecology Reports' journal"; however, after a while, I like it a lot. It could be very helpful for people from other fields. -->

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
<!--JN: I like the second argument above. We can also add that the use of open-source software democratize scientific research (as opposed to the closed+expensive software) -->
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
<!--language is older (1991?), but in 1995 it was officially released for the first time-->
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
<!--JN: should we cite Pebesma2019a as it is a work in progress? -->
<!--JN: also - maybe we should cite http://book.ecosens.org/RSEbook/?-->
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

@@

<!-- JN: terra new vector representations -->

<!-- MH: I probably would not add terra here. As far as I see it, the vector support in terra is basically just support to combine vector with raster data. The main focus of the package is raster data though -->

@@

<!--JN: idea - a figure comparing viz of several packages-->

>>>>

<!-- JN: my first thought was - "do we even need this kind of introduction in the 'Current Landscape Ecology Reports' journal"; however, after a while, I like it a lot. It could be very helpful for people from other fields. -->
<!-- MH: Yeah, I kind of like it. It sets the stage I think. Also, I hope the paper will be read by people that are not really familiar with landscape ecology as a field and just came across some spatial data -->

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratize scientific research as free-of-cost software removes one gatekeeper for researches without access to proprietary software.

One successful example of an open-source project is the *R* programming language, and its extensions called packages \cite{RCoreTeam2019}.
Firstly released in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
<!--JN: language is older (1991?), but in 1995 it was officially released for the first time-->
<!-- MH: Since we write "released" I think it's ok to use 1995? Especially since this was most likely the time a broader audience used it? -->
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006,Lovelace2019}.
The expanding CRAN Task Views document this: *Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} currently list about 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen with the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.
<!--JN: should we cite Pebesma2019a as it is a work in progress? -->
<!-- MH: I think this does not speak against citing it, but also happy to remove it -->
<!--JN: also - maybe we should cite http://book.ecosens.org/RSEbook/?-->
<!-- MH: I dont know it, but sure if it fits -->
A recent overview over the progress of *R* to handle spatial data in general can be found in \cite{Bivand2020}.

@@

<!-- JN: terra new vector representations -->
<!-- MH: I probably would not add terra here. As far as I see it, the vector support in terra is basically just support to combine vector with raster data. The main focus of the package is raster data though -->

@@

<!--JN: idea - a figure comparing viz of several packages-->
<!-- MH: I like the idea: Maybe one base plot, one fancy map (tmap) and a screenshot of an interactive map? -->

### 2020-12-21 paper/paper.tex
:pencil: Address some comments

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.

@@

To better understand how the landscape ecology community uses* R*, we conducted a short survey and used mailing lists and social media to reach the community.
In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
Other, less frequent answers were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig. \ref{fig:fig-survey} D)).

@@

\caption{Results of the online survey about open-source software tools in R for landscape ecology. Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology. 'others' includes all answers with less than five total mentions.}\label{fig:fig-survey}
\end{figure}

>>>>

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, which allows using the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratize scientific research as free-of-cost software removes one gatekeeper for researches without access to proprietary software.

@@

To better understand how the landscape ecology community uses* R*, we conducted a short survey and used mailing lists and social media to reach the community.
In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
Other, less frequent answers were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

Lastly, when asked how useful \emph{R} is currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig. \ref{fig:fig-survey} D)).

@@

\caption{Results of the online survey about open-source software tools in R for landscape ecology. Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology. The 'others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
\end{figure}
